Whittaker Models for Real Groups

### 2024-02-23 whittaker.tex
Significant update to the action of Q(R) on real orbits

\begin{document}
\title{Whittaker Models for Real Groups}
\author{Jeffrey Adams + AA}
\maketitle

@@

Before embarking on the proof, let us point out that the bijections are defined using rather deep results of representation theory. However, we shall prove in Section~\ref{sec:explicit} that all bijections $(i) \leftrightarrow (j)$, for $i,j \in \{(1), \dots, (4)\}$, can be spelled out in elementary terms. Furthermore we shall prove, using elementary arguments, that (3) $\leftrightarrow$ (4) coincides with the Kostant--Sekiguchi correspondence.

@@

\paragraph*{$\bullet$ Action on~(4)} The actions of $Q(\R)$ on (3) and (4) both use an interpretation of $Q(\R)$ in terms of group cohomology. We begin with the action on (4), because the underlying ideas in Galois cohomology may be more familiar to some readers.

We first recall some notation from group cohomology. Let~$\Gamma$ be the Galois group~$\Z/2\Z$ of~$\R$. When~$A$ is a group and $\tau$ is an involutive automorphism of~$A$, we may consider the cohomology sets $H^0_\tau(\Gamma, A)$ and $H^1_\tau(\Gamma, A)$ attached to the action of~$\Gamma$ on~$G$ using~$\tau$.  These are pointed sets, and not groups in general if~$A$ isn't abelian. We may view $H^0_\tau(\Gamma, A)$ as the fixed-point-set $A^{\tau}$, and $H^1_\tau(\Gamma,A)$ as the quotient of $A^{-\tau} = \{ a \in A, a\tau(a)=1\}$ by the equivalence relation~$\sim$ generated by $a \sim x a \tau(x^{-1})$ for all $x \in G$.

Let us now define the action of $Q(\R)$ on~(4). By definition we have the exact sequence $1\rightarrow Z \rightarrow G \rightarrow \Gad\rightarrow 1$. It 
gives rise to a long exact sequence of pointed sets:
\begin{equation} \label{long_ptset}
1\rightarrow Z(\R) \rightarrow G(\R) \rightarrow \Gad(\R) \rightarrow H^1_\sigma(\Gamma,Z)\rightarrow H^1_\sigma(\Gamma,G)\rightarrow H^1_\sigma(\Gamma,\Gad).
\end{equation}
By definition the connecting map  $ \Gad(\R) \rightarrow H^1_\sigma(\Gamma,Z)$ sends $g \in \Gad(\R)$ to the equivalence class of $g \sigma(g^{-1})$ in $H^1_\sigma(\Gamma, Z)$. Going through the definitions we see that   two elements  $g_1, g_2$ have the same image if and only if $g_1 g_{2}\inv \in G(\R)_{\mathrm{ad}}$; therefore the image of the connecting map is in canonical bijection with $Q(\R)=\Gad(\R)/G(\R)_{\mathrm{ad}}$, and by exactness of~\eqref{long_ptset} we get:

\begin{lemma}
  \label{l:Q}
The above discussion determines a canonical bijection 
 \[ Q(\R) \simeq\kernel(H^1_\sigma(\Gamma,Z)\rightarrow H^1_\sigma(\Gamma,G)).\]
\end{lemma}

Using this we define an action of~$Q(\R)$ on $(\Op \cap i\g(\R))/G(\R)$. Suppose~$X \in \g$ is nilpotent. Consider the intersection of the orbit $\Omega_X=\mathrm{Ad}(G) \cdot X$ with $\g(\R)$, and its decomposition into $G(\R)$-orbits. The map 
\begin{align*} (\Omega_{X} \cap \g(\R))/G(\R) & \to \kernel\left(H^1_\sigma(\Gamma,\mathrm{Stab}_G(X))\rightarrow H^1_\sigma(\Gamma,G)\right) \\ \mathrm{Ad}(g) \cdot X & \mapsto \text{class of $g^{-1} \sigma(g)$ in $H^1_\sigma(\Gamma,\mathrm{Stab}_G(X))$} \end{align*} is a bijection: see \cite[Lemma 5.2]{galois}. 
In the case where $X$ lies on the principal nilpotent orbit $\Op$, the stabilizer $\mathrm{Stab}_G(X)$ is the direct product of $Z(G)$ with a unipotent group\footnote{AA 19-II-2024: should we insert a reference?}, which implies $H^1(\Gamma, \mathrm{Stab}_G(X)) =H^1(\Gamma,Z)$. Using Lemma~\ref{l:Q}, we get a bijection $Q(\R) \leftrightarrow (\Op \cap i\g(\R))/G(\R)$, which depends only on the choice of~$X \in \Op$; and a simply transitive action of~$Q(\R)$ on~$(\Op \cap i\g(\R))/G(\R)$, which is entirely canonical.\footnote{AA 19-II-2023: we should say why the map $\pi \mapsto \WF(\pi)$ is equivariant for the action of $Q(\R)$ on both sides.  You wrote this is obvious, but is it really obvious if we choose this description of the action? }

>>>>

\begin{document}
\title{Whittaker Models for Real Groups}
\author{Jeffrey Adams \& Alexandre Afgoustidis}
\maketitle

@@

Before embarking on the proof, let us point out that the maps in the theorem are defined using rather deep results of representation theory. However, we shall prove in Section~\ref{sec:explicit} that all bijections $(i) \leftrightarrow (j)$, for $i,j \in \{(1), \dots, (4)\}$, can be spelled out in elementary terms. Furthermore we shall prove, using elementary arguments, that (3) $\leftrightarrow$ (4) coincides with the Kostant--Sekiguchi correspondence.

@@

\paragraph*{$\bullet$ Action on~(4)} The actions of $Q(\R)$ on (3) and (4) are quite similar, and the proof that they are simply transitive use an interpretation of $Q(\R)$ in terms of group cohomology.

We begin with the case of the wave-front set (4), because the underlying ideas in Galois cohomology may be more familiar to some readers.

To define the action of~$Q(\R)$ on~(4), first observe that~$\Gad$ acts on~$\g$ by the adjoint action. This action preserves~$\Op$, and the restriction to~$\Gad(\R)$ preserves~$\Op\cap (i\g(\R))$. Since elements on a given~$G(\R)_{\mathrm{ad}}$-orbit are in the same~$G(\R)$-orbit, this induces an action of $Q(\R)=\Gad(\R)/G(\R)_{\mathrm{ad}}$ on~$\Op\cap (i\g(\R))/G(\R)$.

\begin{proposition}
\begin{enumerate} 
\item The action of~$Q(\R)$ on $\Op\cap (i\g(\R))/G(\R)$ is simply transitive.
\item The map $\pi \mapsto \WF(\pi)$ is $Q(\R)$-equivariant.
\end{enumerate}
\end{proposition}

The proof of~(1) uses group comology, and we first recall some general notation. Let~$\Gamma$ be the Galois group~$\Z/2\Z$ of~$\R$. When~$A$ is a group and $\tau$ is an involutive automorphism of~$A$, we may consider the cohomology sets $H^0_\tau(\Gamma, A)$ and $H^1_\tau(\Gamma, A)$ attached to the action of~$\Gamma$ on~$G$ using~$\tau$.  These are pointed sets, and not groups in general if~$A$ isn't abelian. We may view $H^0_\tau(\Gamma, A)$ as the fixed-point-set $A^{\tau}$, and $H^1_\tau(\Gamma,A)$ as the quotient of $A^{-\tau} = \{ a \in A, a\tau(a)=1\}$ by the equivalence relation~$\sim$ generated by $a \sim x a \tau(x^{-1})$ for all $x \in G$.

We can use group cohomology to define an action of $Q(\R)$ on~(4). By definition we have the exact sequence $1\rightarrow Z \rightarrow G \rightarrow \Gad\rightarrow 1$. It 
gives rise to a long exact sequence of pointed sets:
\begin{equation} \label{long_ptset}
1\rightarrow Z(\R) \rightarrow G(\R) \rightarrow \Gad(\R) \rightarrow H^1_\sigma(\Gamma,Z)\rightarrow H^1_\sigma(\Gamma,G)\rightarrow H^1_\sigma(\Gamma,\Gad).
\end{equation}
By definition the connecting map  $ \Gad(\R) \rightarrow H^1_\sigma(\Gamma,Z)$ sends $g \in \Gad(\R)$ to the equivalence class of $g \sigma(g^{-1})$ in $H^1_\sigma(\Gamma, Z)$. Going through the definitions we see that   two elements  $g_1, g_2$ have the same image if and only if $g_1 g_{2}\inv \in G(\R)_{\mathrm{ad}}$; therefore the image of the connecting map is in canonical bijection with $Q(\R)=\Gad(\R)/G(\R)_{\mathrm{ad}}$, and by exactness of~\eqref{long_ptset} we get:

\begin{lemma}
  \label{l:Q}
The above discussion determines a canonical bijection 
 \begin{equation}\label{bij_q_1} Q(\R) \leftrightarrow \kernel(H^1_\sigma(\Gamma,Z)\rightarrow H^1_\sigma(\Gamma,G)).\end{equation}
\end{lemma}

Using this we define an action of~$Q(\R)$ on $(\Op \cap i\g(\R))/G(\R)$. Suppose~$X \in \g$ is nilpotent. Consider the intersection of the orbit $\Omega_X=\mathrm{Ad}(G) \cdot X$ with $\g(\R)$, and its decomposition into $G(\R)$-orbits. The map 
\begin{align} \label{bij_q_2}\beta_X\colon (\Omega_{X} \cap \g(\R))/G(\R) & \to \kernel\left(H^1_\sigma(\Gamma,\mathrm{Stab}_G(X))\rightarrow H^1_\sigma(\Gamma,G)\right) \\ \mathrm{Ad}(g) \cdot X & \mapsto \text{class of $ g\sigma(g^{-1})$ in $H^1_\sigma(\Gamma,\mathrm{Stab}_G(X))$} \nonumber \end{align} 
is a bijection: see \cite[Lemma 5.2]{galois}. 
In the case where $X$ lies on the principal nilpotent orbit $\Op$, the stabilizer $\mathrm{Stab}_G(X)$ is the direct product of $Z(G)$ with a unipotent group\footnote{AA 19-II-2024: should we insert a reference?}, which implies $H^1(\Gamma, \mathrm{Stab}_G(X)) =H^1(\Gamma,Z)$. Using Lemma~\ref{l:Q}, we get a bijection 
\[\Delta_X\colon (\Op \cap i\g(\R))/G(\R) \leftrightarrow Q(\R).\]
This bijection depends  on the choice of~$X \in \Op$; but it defines a simply transitive action of~$Q(\R)$ on~$(\Op \cap i\g(\R))/G(\R)$, which is independent of the choice of~$X$.

Let us now prove the Proposition.

First we prove that~$Q(\R)$ acts freely on the set a  $(\Op \cap i\g(\R))/G(\R)$ of orbits. Let~$\omega$ be an orbit, and let~$X$ be a point of~$\omega$. If~$q$ is an element of~$Q(\R)$, then going through the definitions of~\eqref{bij_q_1}--\eqref{bij_q_2} we see that~$\Delta_X(q \cdot \omega) = q$; therefore if~$q\cdot \omega=\omega$, then~$q=1$.

Next we prove that the action is transitive. Let~$\omega, \omega'$ be $G(\R)$-orbits in  $(\Op \cap i\g(\R))$. Fix an element~$X$ of~$\omega$ and consider the bijections~$\beta_X$ and~$\Delta_X$ above. Let~$q \in Q(\R)$ be the element~$\Delta_X(\omega')$. We have seen that $q = \Delta_X(q \cdot \omega)$, therefore \mbox{$\Delta_X(q \cdot \omega)=\Delta_X(\omega')$}, which proves~$q \cdot \omega = \omega'$ and concludes the proof of~(1).

For the proof of part~(2) of the proposition, use the asymptotic cone.
